---
title: "PH 718 Data Management and Visualization in `R`"
subtitle: "Part 3: Basic Graphics in `R`"
author: 
  - Zhiyang Zhou (zhou67@uwm.edu, [zhiyanggeezhou.github.io](https://zhiyanggeezhou.github.io/))
date: "`r format(Sys.time(), '%Y/%m/%d %H:%M:%S')`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

## `plot()`

-   Primary way to plot data in base `R`.

-   `plot(x, y)` produces a scatterplot of the numbers in `x` vs. the
    numbers in `y`.

-   Additional options available. For example, passing in the argument
    `xlab` will result in a label on the x-axis. To find out more
    information about the `plot()` function, type `?plot`.

```{r chunk1, echo=F,include = T, eval = T, cache = T, massage = T}
set.seed(718) # to make the subsequent randomization reproducible
x <- rnorm(100) # yield 100 random numbers following N(0,1)
y <- rnorm(100)
plot(x, y, 
     xlab = "this is the x-axis", 
     ylab = "this is the y-axis", 
     main = "Plot of X",
     col = 'red',
     pch = 'X',
     cex = 1
    )
```

## `contour()`

-   Producing a *contour plot* in order to represent three-dimensional
    data, like a topographical map. It takes three arguments:
    -   a vector of `x` values (the first dimension),
    -   a vector of `y` values (the second dimension), and
    -   a matrix whose elements correspond to the `z` value (the third
        dimension) for each pair of (`x`, `y`) coordinates.

```{r chunk2, echo=F,include = T, eval = F, cache = T, massage = T}
x <- seq(-pi, pi, length = 50)
y <- x
f <- outer(x, y, function(x, y){x^2+y^2}) # outer product of x and y
contour(x, y, f)
```

## `image()`

-   Working the same way as `contour()`, and producing a *heatmap* (a
    color-coded plot whose colors depend on the `z` value).

```{r chunk3, echo=F,include = T, eval = F, cache = T, massage = T}
x <- seq(-pi, pi, length = 50)
y <- x
f <- outer(x, y, function(x, y){x^2+y^2}) # outer product of x and y
image(x,y,f)
```

## `persp()`

-   Producing a three-dimensional plot.

```{r chunk4, echo=F,include = T, eval = F, cache = T, massage = T}
x <- seq(-pi, pi, length = 50)
y <- x
f <- outer(x, y, function(x, y){x^3+y^2}) # outer product of x and y
persp(x, y, f)
persp(x, y, f, theta = 30) # theta and phi control the angle at which the plot is viewed
persp(x, y, f, theta = 30, phi = 20)
persp(x, y, f, theta = 30, phi = 70)
persp(x, y, f, theta = 30, phi = 40)
```

## A toy example (ISL, pp. 54)

-   The `College` data set can be found on the book website
    (<https://www.statlearning.com/s/College.csv>). It contains a number
    of variables for 777 different universities and colleges in the US.
    The variables are
    -   `Private`: Public/private indicator
    -   `Apps`: Number of applications received
    -   `Accept`: Number of applicants accepted
    -   `Enroll`: Number of new students enrolled
    -   `Top10perc`: New students from top 10% of high school class
    -   `Top25perc`: New students from top 25% of high school class
    -   `F.Undergrad`: Number of full-time undergraduates
    -   `P.Undergrad`: Number of part-time undergraduates
    -   `Outstate`: Out-of-state tuition
    -   `Room.Board`: Room and board costs
    -   `Books`: Estimated book costs
    -   `Personal`: Estimated personal spending
    -   `PhD`: Percent of faculty with Ph.D. degree
    -   `Terminal`: Percent of faculty with terminal degree
    -   `S.F.Ratio`: Student/faculty ratio
    -   `perc.alumni`: Percent of alumni who donate
    -   `Expend`: Instructional expenditure per student
    -   `Grad.Rate`: Graduation rate
-   Use the `read.csv()` function to read the data into `R`. Call the
    loaded data `college`.

```{r chunk5, echo=F,include = T, eval = F, cache = T, massage = T}
college = read.csv(file = "https://www.statlearning.com/s/College.csv")
```

-   The first column is the name of each university. We don't really
    want `R` to treat this as a variable. Instead, it is better to have
    `row.names` recording these names before removing them. Try the
    following commands:

```{r chunk6, echo=F,include = T, eval = F, cache = T, massage = T}
rownames(college) <- college[, 1]
college <- college[, -1]
```

-   Use `summary()` function to produce a numerical summary of variables
    in the data set.

```{r chunk7, echo=F,include = T, eval = F, cache = T, massage = T}
college$Private = as.factor(college$Private)
summary(college)
```

-   Use `pairs()` function to produce a scatterplot matrix of the first
    ten variables of the data.

```{r chunk8, echo=F,include = T, eval = F, cache = T, massage = T}
pairs(college[,1:10])
```

-   Use `boxplot()` function to produce side-by-side boxplots of
    `Outstate` versus `Private`.

```{r chunk9, echo=F,include = T, eval = F, cache = T, massage = T}
boxplot(Grad.Rate ~ Private, data = college)
```

-   Create a new qualitative variable, called `Elite`, by dividing
    universities into two groups based on whether or not the proportion
    of students coming from the top 10% of their high school classes
    exceeds 50%.

```{r chunk10, echo=F,include = T, eval = F, cache = T, massage = T}
Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite) # or college$Elite = Elite
```

-   Use `summary()` function to see how many elite universities there
    are.

```{r chunk11, echo=F,include = T, eval = F, cache = T, massage = T}
summary(college$Elite)
```

-   Use `boxplot()` function to produce side-by-side boxplots of
    `Outstate` vs. `Elite`.

```{r chunk12, echo=F,include = T, eval = F, cache = T, massage = T}
boxplot(Outstate ~ Elite, data = college)
```

-   Use `hist()` function to produce histograms with different numbers
    of bins for quantitative variables. You may find the command
    `par(mfrow = c(2, 2))` useful: it will divide the print window into
    four regions so that four plots can be made simultaneously.

```{r chunk13, echo=F,include = T, eval = F, cache = T, massage = T}
par(mfrow = c(1, 2))
hist(college$Top25perc, breaks = 1, freq = T, main = 'breaks=1')
hist(college$Top25perc, breaks = 10, freq = T, main = 'breaks=10')
hist(college$Top25perc, breaks = 100, freq = T, main = 'breaks=100')
hist(college$Top25perc, breaks = 1000, freq = T, main = 'breaks=1000')
par(mfrow = c(1, 1))
```
