---
title: "PH 718 Data Management and Visualization in `R`"
subtitle: "Part 4: Introduction to `Tidyverse`"
author: 
  - Zhiyang Zhou (zhou67@uwm.edu, [zhiyanggeezhou.github.io](https://zhiyanggeezhou.github.io/))
date: "`r format(Sys.time(), '%Y/%m/%d %H:%M:%S')`"
output: pdf_document
---

## What is the `Tidyverse` (https://www.tidyverse.org/)

- A collection of `R` packages designed for data science. 
These packages share a common philosophy, grammar, and data structure, making data manipulation, visualization, and modeling easier and more efficient.

| Package   | Purpose                                               |
|-----------|-------------------------------------------------------|
| `ggplot2` | Data visualization                                    |
| `dplyr`   | Data manipulation (filter, summarize, mutate, etc.)   |
| `tidyr`   | Data tidying (reshaping data)                         |
| `readr`   | Reading CSVs and text files                           |
| `tibble`  | Modern replacement for data frames                    |
| `purrr`   | Functional programming (working with lists)           |
| `stringr` | String manipulation                                   |
| `forcats` | Handling categorical data (factors)                   |

## Why use `Tidyverse`?

- Consistency: Functions follow similar syntax and principles.
- Readability: Code is easier to write and understand.
- Efficiency: Optimized for modern data workflows.
- Integration: `Tidyverse` functions work well together.

## Installing and loading `Tidyverse`

```{r, eval = F}
install.packages("tidyverse")
library(tidyverse) # load ALL core tidyverse packages
```

If you need a specific package, you can load it individually, e.g.,

```{r, eval = F}
library(dplyr)
```

--- 

## Toy examples illustrating features of `Tidyverse` 

### Data manipulation with `dplyr`

```{r, eval = F}
mtcars_filtered1 = 
  mtcars %>%
  filter(cyl == 6) %>%       # Keep only cars with 6 cylinders
  select(mpg, hp, wt) %>%    # Select relevant columns
  arrange(desc(mpg))         # Arrange in descending order of mpg
```

or, equivalently, without using functions in `Tidyverse`,

```{r, eval = F}
mtcars_filtered3 <- mtcars[mtcars$cyl == 6, ]
mtcars_filtered3 <- mtcars_filtered3[, c("mpg", "hp", "wt")]
mtcars_filtered3 <- mtcars_filtered3[order(-mtcars_filtered3$mpg), ]
```

### Creating elegant, layered plots with `ggplot2`

```{r, eval = F}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) + # aesthetics
  geom_point() + # type of plot
  theme_minimal() # style
```

or, similarly, without using `ggplot2`

```{r, eval = F}
colors <- c("red", "blue", "green") 
cyl_factor <- as.factor(mtcars$cyl)
color_map <- colors[as.numeric(cyl_factor)]
plot(mtcars$wt, mtcars$mpg, col = color_map, pch = 16,
     xlab = "Weight (wt)", ylab = "Miles per Gallon (mpg)",
     main = "Scatter plot of wt vs mpg by cyl")
legend("topright", legend = levels(cyl_factor), col = colors, pch = 16, title = "Cylinders")
```

### Reshaping data with `tidyr`

```{r, eval = F}
# Converting wide data to the long format
wide_data <- data.frame(
  name = c("Alice", "Bob", "Cindy"), 
  math = c(90, 85, 69), 
  science = c(88, 92, 99)
)
long_data1 <- wide_data %>%
  pivot_longer(cols = math:science, names_to = "subject", values_to = "score")
```

or, equivalently, without using functions in `Tidyverse`,

```{r, eval = F}
# Converting wide data to the long format
wide_data <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(23, 25, 22),
  score = c(90, 85, 88)
)
long_data2 <- reshape(
  wide_data, 
  varying = c("math", "science"), 
  v.names = "score", 
  timevar = "subject", 
  times = c("math", "science"), 
  direction = "long"
)
rownames(long_data2) <- NULL
long_data2 = subset(long_data2, select = -id)
```

---

## Tibble

### What is a tibble?

- A modern version of data frame optimized for better readability and usability in `R`
- Working seamlessly with packages belonging to `tidyverse` 
- Rumor: "Tibble" is similar to how New Zealanders pronounce "table", and Dr. Hadley Wickham, the creator of `tibble`, is from New Zealand.

### Creating tibbles

```{r, eval = F}
# Creating a tibble manually with tibble::tibble()
students1 <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  age = c(23, 25, 22),
  score = c(90, 85, 88)
)
students1
```

```{r, eval = F}
# For small datasets, tibble::tribble() allows for a more intuitive way
students2 <- tribble(
  ~name,    ~age, ~score,
  "Alice",    23,    90,
  "Bob",      25,    85,
  "Charlie",  22,    88
)
students2
```

```{r, eval = F}
# Converting between tibbles and data frames
iris
iris_t <- as_tibble(iris)
iris_t 
iris_df = as.data.frame(iris_t)
iris_df
```

## Accessing subsets of a tibble

```{r, eval = F}
iris_t$Species     
iris_t[,c("Petal.Length","Petal.Width","Species")]
iris_t[iris_t$Species=='setosa', ]
subset(iris_t, select = c(Petal.Length,Petal.Width,Species))
subset(iris_t, select = -c(Petal.Length,Petal.Width,Species))
```

### Key advantages of tibbles

```{r, eval = F}
# Tibbles do not convert character vectors into factors automatically
df <- data.frame(name = c("Alice", "Bob"), age = c(23, 25))
tib <- tibble(name = c("Alice", "Bob"), age = c(23, 25))

str(df)   # name is converted to factor
str(tib)  # name remains as character
```

```{r, eval = F}
# Tibbles print only the first 10 rows and all the columns that fit on screen
large_tibble <- tibble(x = 1:1000, y = runif(1000))
large_tibble
```

---

## Pipe operator `%>%`

### What is the pipe operator

- A fundamental feature in `tidyverse`
- Designed to improve code readability and workflow efficiency
  - Chaining multiple operations together in a logical sequence
  - Making data manipulation in `R` clearer and more intuitive

### History of pipes

- 1970s: Pipe `|` was introduced in Unix, allowing command-line programs to pass output to the next command;
- Languages like F#, Elixir, and Haskell have similar piping mechanisms;
- 2014: `%>%` was introduced in `R` package `magrittr`;
- 2016: `%>%` was integrated into `tidyverse`;
- 2021: `|>` was introduced in base R 4.1 and an alternative for users who prefer not to load `tidyverse`.

### Why use the pipe operator

```{r, eval = F}
# Sample dataset
student_data <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
  score = c(85, 92, 78, NA, 95, 88)
)
```

- Given the above dataset of student exam scores, we want to:
  1. Remove any students with missing scores (NA).
  1. Standardize the scores.
  1. Screen out students who scored below average after standardization.
  1. Sort the remaining students by their standardized scores in descending order.

```{r, eval = F}
# If we use multiple intermediate variables
clean_data <- na.omit(student_data)
mean_score <- mean(clean_data$score)
sd_score <- sd(clean_data$score)
clean_data$z_score <- (clean_data$score - mean_score) / sd_score
filtered_data <- clean_data[clean_data$z_score > 0, ]
sorted_data <- filtered_data[order(-filtered_data$z_score), ]
sorted_data

## Problem: requires multiple intermediate variables, making the workflow less readable.
```

```{r, eval = F}
# If we use a nested function
subset(
  transform(
    na.omit(student_data),
    z_score = (score - mean(score)) / sd(score)
  ),
  z_score > 0
)[order(-subset(
  transform(
    na.omit(student_data),
    z_score = (score - mean(score)) / sd(score)
  ),
  z_score > 0
)$z_score), ]

## Problem: reversed order of operations (we must read inside-out), making the workflow less readable
```

```{r, eval = F}
# If we use %>%
student_data %>%
  na.omit() %>%  # Step 1: Remove missing values
  transform(z_score = (score - mean(score)) / sd(score)) %>%  # Step 2: standardize scores
  subset(z_score > 0) %>%  # Step 3: Keep students with above-average scores
  .[order(-.$z_score), ] # Step 4: Sort in descending order

## More readable, following a left-to-right logic, no need for temporary variables, and easier to debug
```

### How the pipe operator works

- Passing the output of one function as the first argument to the next function

```{r, eval = F}
mean(mtcars$mpg)
# equiv.
mtcars$mpg %>% mean()
```

```{r, eval = F}
head(iris)
# equiv.
iris %>% head()
```

```{r, eval = F}
x <- runif(10)
round(exp(diff(log(x))), 1)
# equiv.
x %>% log() %>% diff() %>% exp() %>% round(1)
```

- Passing a dataset to positions where placeholders (dots) lie
  - The dot is not a general placeholder for any function argument
  but specifically represents the dataset in the pipeline;
  it is only necessary when the dataset is not automatically placed as the first argument in a function

```{r, eval = F}
student_data <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
  score = c(85, 92, 78, NA, 95, 88)
)
student_data %>%
  na.omit() %>%
  transform(z_score = (score - mean(score)) / sd(score)) %>%
  subset(z_score > 0) %>%
  .[order(-.$z_score), ] # R is going to replace the two dots with the previously generated dataset
```

---

## `tidyr`

---

## `dplyr`

---

## `ggplot2`