---
title: "PH 718 Data Management and Visualization in `R`"
subtitle: "Part 2: How to Use R Markdown"
author: 
  - Zhiyang Zhou (zhou67@uwm.edu, [zhiyanggeezhou.github.io](https://zhiyanggeezhou.github.io/))
date: "`r Sys.Date()`"
output: pdf_document
bibliography: UWM_PH718_2025Spring_Part02.bib
link-citations: false
---

# Introduction

R Markdown is versatile tools for creating documents, presentations, dashboards, and more.

This guide covers basic Markdown syntax, features of R Markdown, and examples to help you get started.

---

# Installation of R Markdown

You can install the R Markdown package from CRAN with:
```{r eval=FALSE}
install.packages("rmarkdown")
```
The following link provides a quick tour of R Markdown:
https://rmarkdown.rstudio.com/lesson-1.html

---

# Basics of Markdown

Markdown is a lightweight language used for formatting text. 
It is simple to use and widely supported. Here are some basic Markdown features:

### Headings: using `#`

- `# Heading 1`
- `## Heading 2`
- `### Heading 3`

### Emphasis

- **Bold:** `**bold text**`
- *Italics:* `*italicized text*`
- ~~Strikethrough:~~ `~~strikethrough~~`

### Lists

* **Unordered list:**
  - Use `-` or `*` for items.
    - Nested list item.
* **Ordered list:**
  1. First item
  1. Second item
  1. Third item

### Links and images

- **Links:** `[text](URL)`
  - Example: [R Markdown website](https://rmarkdown.rstudio.com/)
- **Images:** `![alt text](image_URL)`
  - Example: `![Sample Image](https://via.placeholder.com/150)`

### Code blocks

- Inline code: Use backticks \`code\`.
- Fenced code block:
  ```
  print("Hello, Markdown!")
  ```
- Blockquote: `> blockquote`


### Extended syntax 

https://www.markdownguide.org/cheat-sheet/

---

# Basics of R Markdown

### R code chunks

R Markdown extends Markdown by allowing you to embed R code within your document. A chunk starts with three backticks and `{r}`.

Example:
```{r, cache=T}
# A simple R code chunk
data(cars)
summary(cars)
```

```{r, cache=T}
# A simple R code chunk
result1 = 0 
for (i in 1:10^4){
  result1 = result1 + i^2
}
result1
```


### Inline code

You can include R code inline using backticks and `r`:
`r 2 + 2` becomes `r 2 + 2`.

---

# Additional features

### Plots
Include plots using R code chunks:
```{r, cache=T}
# Example of a scatter plot
data(mtcars)
plot(
  mtcars$mpg, 
  mtcars$wt, 
  main = "Scatter Plot of MPG vs Weight", 
  xlab = "Weight", 
  ylab = "MPG"
)
```

### Tables
Use functions like `knitr::kable()` or `gt` for formatted tables:
```{r, cache=T}
library(knitr)
kable(head(mtcars), caption = "Sample Table of mtcars Dataset")
```

### Caching
Speed up document rendering with caching:
```{r, cache=T}
# Cached code chunk
summary(mtcars)
```

---

# Incorporating references using a BibTeX (`.bib`) file 

## Citing the works of others in your writing is essential

- Acknowledgment of original ideas: 
citations give proper credit to the original authors for their contributions, recognizing their intellectual property and efforts.

- Upholding academic integrity: Citing sources is crucial to avoid plagiarism, 
which is the unethical practice of presenting someone else's work or ideas as your own.

- Demonstrating research depth: By referencing existing literature, 
you showcase the breadth and depth of your research, 
indicating a comprehensive understanding of the subject matter.

- Supporting your arguments: Citations provide evidence for your claims, 
strengthening your arguments by linking them to established knowledge. 

- Facilitating further research: Proper citations allow readers to trace the original sources, enabling them to explore the topic further and verify information.

## Create a BibTeX (`.bib`) file:

Begin by creating a `.bib` file that contains all your bibliographic entries. 
Each entry should follow the BibTeX format.

### Templates for different types of entries

An article from a magazine or a journal. For example:

```{eval=FALSE}
@article{greenwade93,
    author  = "George D. Greenwade",
    title   = "The {C}omprehensive {T}ex {A}rchive {N}etwork ({CTAN})",
    year    = "1993",
    journal = "TUGBoat",
    volume  = "14",
    number  = "3",
    pages   = "342--351"
}
```

A published book. For example:

```{eval=FALSE}
@book{goossens93,
    author    = "Michel Goossens and Frank Mittelbach and Alexander Samarin",
    title     = "The LaTeX Companion",
    year      = "1993",
    publisher = "Addison-Wesley",
    address   = "Reading, Massachusetts"
}
```

An article accepted by a conference. For example:

```{eval=FALSE}
@inproceedings{lesk:1977,
  title={Computer Typesetting of Technical Journals on {UNIX}},
  author={Michael Lesk and Brian Kernighan},
  booktitle={Proceedings of American Federation of
             Information Processing Societies: 1977
             National Computer Conference},
  pages={879--888},
  year={1977},
  address={Dallas, Texas}
}
```

You can find more information about other entry types
[here](https://en.wikibooks.org/wiki/LaTeX/Bibliography_Management#BibTeX).

## Link the `.bib` file to your `.Rmd` file

In the header of your `.Rmd` file, 
specify the path to your bibliography file.
If the `.bib` and `.Rmd` files are in the same folder, 
you may merely specify the name of the `.bib` file 
rather than the full path to it.

```{eval=FALSE}
bibliography: UWM_PH718_2025Spring_Part02.bib
```

## Cite references

Use the `@` symbol followed by the citation key.

### In-text citation: `@Wickham2016`

As demonstrated in @Wickham2016, data visualization is a powerful tool.

### Parenthetical citation: `[@Wickham2016]`

Data visualization is a powerful tool [@Wickham2016].

### For multiple citations, separate the keys with semicolons

Several `R` packages [@tidyr2024; @R2024] are fairly useful.

## Generate the bibliography

The cited references are automatically compiled into a bibliography section 
at the end of your document, 
even if you haven't explicitly added a section header.

---

# Resources

- [Markdown Guide](https://www.markdownguide.org/)
- [R Markdown Documentation](https://rmarkdown.rstudio.com/)

# References