---
title: "How to Use R Markdown"
author: "Your Name"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction

R Markdown is versatile tools for creating documents, presentations, dashboards, and more. 
This guide covers basic Markdown syntax, features of R Markdown, and examples to help you get started.

---

# Installation of R Markdown

You can install the R Markdown package from CRAN with:
```{r eval=FALSE}
install.packages("rmarkdown")
```
The following link provides a quick tour of R Markdown:
https://rmarkdown.rstudio.com/lesson-1.html

---

# Basics of Markdown

Markdown is a lightweight language used for formatting text. 
It is simple to use and widely supported. Here are some basic Markdown features:

### Headings: using `#`

- `# Heading 1`
- `## Heading 2`
- `### Heading 3`

### Emphasis

- **Bold:** `**bold text**`
- *Italics:* `*italicized text*`
- ~~Strikethrough:~~ `~~strikethrough~~`

### Lists

- **Unordered list:**
  - Use `-` or `*` for items.
    - Nested list item.
- **Ordered list:**
  1. First item
  2. Second item
  3. Third item

### Links and images

- **Links:** `[text](URL)`
  - Example: [R Markdown website](https://rmarkdown.rstudio.com/)
- **Images:** `![alt text](image_URL)`
  - Example: `![Sample Image](https://via.placeholder.com/150)`

### Code blocks

- Inline code: Use backticks \`code\`.
- Fenced code block:
  ```
  print("Hello, Markdown!")
  ```
- Blockquote: `> blockquote`


### Extended syntax 

https://www.markdownguide.org/cheat-sheet/

---

# Basics of R Markdown

### R code chunks

R Markdown extends Markdown by allowing you to embed R code within your document. A chunk starts with three backticks and `{r}`.

Example:
```{r}
# A simple R code chunk
summary(cars)
```

### Inline code

You can include R code inline using backticks and `r`:
`r 2 + 2` becomes `r 2 + 2`.

---

# Additional features

### Plots
Include plots using R code chunks:
```{r}
# Example of a scatter plot
plot(mtcars$mpg, mtcars$wt, main = "Scatter Plot of MPG vs Weight", xlab = "Weight", ylab = "MPG")
```

### Tables
Use functions like `knitr::kable()` or `gt` for formatted tables:
```{r}
library(knitr)
kable(head(mtcars), caption = "Sample Table of mtcars Dataset")
```

### Caching
Speed up document rendering with caching:
```{r cache = TRUE}
# Cached code chunk
summary(mtcars)
```

---

# Resources

- [Markdown Guide](https://www.markdownguide.org/)
- [R Markdown Documentation](https://rmarkdown.rstudio.com/)

