---
title: "PH 718 Data Management and Visualization in `R`"
author: "Zhiyang Zhou (zhou67@uwm.edu, [zhiyanggeezhou.github.io](https://zhiyanggeezhou.github.io/))"
date: "`r format(Sys.time(), '%Y/%m/%d %H:%M:%S')`"
output: pdf_document
subtitle: 'Part 4: Introduction to `Tidyverse`'
---

## What is the `Tidyverse` (<https://www.tidyverse.org/>)

A collection of `R` packages designed for data science. 
These packages share a common philosophy, grammar, and data structure, making data manipulation, visualization, and modeling easier and more efficient.
By the way, you may find below an R Markdown table simply created via dashes and bars. 
It could be more convenient to resort to an online tool: <https://www.tablesgenerator.com/markdown_tables>

| Package   | Purpose                                     |
|-----------|---------------------------------------------|
| `ggplot2` | Data visualization                          |
| `dplyr`   | Data manipulation                           |
| `tidyr`   | Rreshaping data                             |
| `readr`   | Reading CSV and text files                  |
| `tibble`  | Modern replacement for data frames          |
| `purrr`   | Functional programming (working with lists) |
| `stringr` | String manipulation                         |
| `forcats` | Handling categorical data (factors)         |

Alternatively,
one may utilize the `knitr::kable` to output the same table with an indexed caption.

```{r, echo=FALSE, results="asis"}
# Create the table as a data frame
tidyr_functions <- data.frame(
  Package = c(
    "`ggplot2`", "`dplyr`", "`tidyr`", "`readr`", "`tibble`", "`purrr`", "`stringr`", "`forcats`"
  ),
  Purpose = c(
    "Data visualization",
    "Data manipulation",
    "Rreshaping data",
    "Reading CSV and text files",
    "Modern replacement for data frames",
    "Functional programming (working with lists)",
    "String manipulation",
    "Handling categorical data (factors)"
  )
)

# Print the table using knitr::kable() for nice formatting
knitr::kable(tidyr_functions, caption = "Summary of Key `tidyr` Functions")
```

## Why use `Tidyverse`

-   Consistency: Functions follow similar syntax and principles.
-   Readability: Code is easier to write and understand.
-   Efficiency: Optimized for modern data workflows.
-   Integration: `Tidyverse` functions work well together.

## Installing and loading `Tidyverse`

```{r, eval = F}
install.packages("tidyverse")
library(tidyverse) # load ALL core tidyverse packages
```

If you need a specific package, you can load it individually, e.g.,

```{r, eval = F}
library(dplyr)
```

---

## Toy examples illustrating features of `Tidyverse`

### Data manipulation with `dplyr`

```{r, eval = F}
mtcars_filtered1 = 
  mtcars %>%
  filter(cyl == 6) %>%       # Keep only cars with 6 cylinders
  select(mpg, hp, wt) %>%    # Select relevant columns
  arrange(desc(mpg))         # Arrange in descending order of mpg
```

or, equivalently, without using functions in `Tidyverse`,

```{r, eval = F}
mtcars_filtered3 <- mtcars[mtcars$cyl == 6, ]
mtcars_filtered3 <- mtcars_filtered3[, c("mpg", "hp", "wt")]
mtcars_filtered3 <- mtcars_filtered3[order(-mtcars_filtered3$mpg), ]
```

### Creating elegant, layered plots with `ggplot2`

```{r, eval = F}
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point() +
  theme_minimal()
```

or, similarly, without using `ggplot2`

```{r, eval = F}
colors <- c("red", "blue", "green") 
cyl_factor <- as.factor(mtcars$cyl)
color_map <- colors[as.numeric(cyl_factor)]
plot(mtcars$wt, mtcars$mpg, col = color_map, pch = 16,
     xlab = "Weight (wt)", ylab = "Miles per Gallon (mpg)",
     main = "Scatter plot of wt vs mpg by cyl")
legend("topright", legend = levels(cyl_factor), col = colors, pch = 16, title = "Cylinders")
```

### Reshaping data with `tidyr`

```{r, eval = F}
# Converting wide data to the long format
wide_data <- data.frame(
  name = c("Alice", "Bob", "Cindy"), 
  math = c(90, 85, 69), 
  science = c(88, 92, 99)
)
long_data1 <- wide_data %>%
  pivot_longer(cols = math:science, names_to = "subject", values_to = "score")
```

or, equivalently, without using functions in `Tidyverse`,

```{r, eval = F}
# Converting wide data to the long format
wide_data <- data.frame(
  name = c("Alice", "Bob", "Cindy"), 
  math = c(90, 85, 69), 
  science = c(88, 92, 99)
)
long_data2 <- reshape(
  wide_data, 
  varying = c("math", "science"), 
  v.names = "score", 
  timevar = "subject", 
  times = c("math", "science"), 
  direction = "long"
)
rownames(long_data2) <- NULL
long_data2 = subset(long_data2, select = -id)
```

---

## Tibble

### What is a tibble?

-   A modern version of data frame optimized for better readability and usability in `R`
-   Working seamlessly with packages belonging to `tidyverse`
-   Rumor: "Tibble" is similar to how New Zealanders pronounce "table", and Dr. Hadley Wickham, the creator of `tibble`, is from New Zealand.

### Creating tibbles

```{r, eval = F}
# Creating a tibble manually with tibble::tibble()
students1 <- tibble(
  name = c("Alice", "Bob", "Charlie"),
  age = c(23, 25, 22),
  score = c(90, 85, 88)
)
students1
```

```{r, eval = F}
# For small datasets, tibble::tribble() allows for a more intuitive way
students2 <- tribble(
  ~name,    ~age, ~score,
  "Alice",    23,    90,
  "Bob",      25,    85,
  "Charlie",  22,    88
)
students2
```

```{r, eval = F}
# Converting between tibbles and data frames
iris
iris_t <- as_tibble(iris)
iris_t 
iris_df = as.data.frame(iris_t)
iris_df
```

## Accessing subsets of a tibble

```{r, eval = F}
iris_t$Species     
iris_t[,c("Petal.Length","Petal.Width","Species")]
iris_t[iris_t$Species=='setosa', ]
subset(iris_t, select = c(Petal.Length,Petal.Width,Species))
subset(iris_t, select = -c(Petal.Length,Petal.Width,Species))
```

### Key advantages of tibbles

```{r, eval = F}
# Tibbles do not convert character vectors into factors automatically
df <- data.frame(name = c("Alice", "Bob"), age = c(23, 25))
tib <- tibble(name = c("Alice", "Bob"), age = c(23, 25))

str(df)   # name is converted to factor
str(tib)  # name remains as character
```

```{r, eval = F}
# Tibbles print only the first 10 rows and all the columns that fit on screen
large_tibble <- tibble(x = 1:1000, y = runif(1000))
large_tibble
```

---

## Pipe operator `%>%`

### What is the pipe operator

-   A fundamental feature in `tidyverse`
-   Designed to improve code readability and workflow efficiency
    -   Chaining multiple operations together in a logical sequence
    -   Making data manipulation in `R` clearer and more intuitive

### History of piping

-   1970s: Pipe `|` was introduced in Unix, allowing command-line programs to pass output to the next command;
-   Languages like F#, Elixir, and Haskell have similar piping mechanisms;
-   2014: `%>%` was introduced in `R` package `magrittr`;
-   2016: `%>%` was integrated into `tidyverse`;
-   2021: `|>` was introduced in base R 4.1 and an alternative for users who prefer not to load `tidyverse`.

### Why use the pipe operator

```{r, eval = F}
# Sample dataset
student_data <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
  score = c(85, 92, 78, NA, 95, 88)
)
```

-   Given the above dataset of student exam scores, we want to:
    1.  Remove any students with missing scores (NA).
    2.  Standardize the scores.
    3.  Screen out students who scored below average after standardization.
    4.  Sort the remaining students by their standardized scores in descending order.

```{r, eval = F}
# If we use multiple intermediate variables
clean_data <- na.omit(student_data)
mean_score <- mean(clean_data$score)
sd_score <- sd(clean_data$score)
clean_data$z_score <- (clean_data$score - mean_score) / sd_score
filtered_data <- clean_data[clean_data$z_score > 0, ]
sorted_data <- filtered_data[order(-filtered_data$z_score), ]
sorted_data

## Problem: requires multiple intermediate variables, making the workflow less readable.
```

```{r, eval = F}
# If we use a nested function
subset(
  transform(
    na.omit(student_data),
    z_score = (score - mean(score)) / sd(score)
  ),
  z_score > 0
)[order(-subset(
  transform(
    na.omit(student_data),
    z_score = (score - mean(score)) / sd(score)
  ),
  z_score > 0
)$z_score), ]

## Problem: reversed order of operations (we must read inside-out), making the workflow less readable
```

```{r, eval = F}
# If we use %>%
student_data %>%
  na.omit() %>%  # Step 1: Remove missing values
  transform(z_score = (score - mean(score)) / sd(score)) %>%  # Step 2: standardize scores
  subset(z_score > 0) %>%  # Step 3: Keep students with above-average scores
  .[order(-.$z_score), ] # Step 4: Sort in descending order

## More readable, following a left-to-right logic, no need for temporary variables, and easier to debug
```

### How the pipe operator works

-   Passing the output of one function as the first argument to the next function

```{r, eval = F}
mean(mtcars$mpg)
# equiv.
mtcars$mpg %>% mean()
```

```{r, eval = F}
head(iris)
# equiv.
iris %>% head()
```

```{r, eval = F}
x <- runif(10)
round(exp(diff(log(x))), 1)
# equiv.
x %>% log() %>% diff() %>% exp() %>% round(1)
```

-   Passing a dataset to positions where placeholders (dots) lie
    -   The dot is not a general placeholder for any function argument but specifically represents the dataset in the pipeline; it is only necessary when the dataset is not automatically placed as the first argument in a function

```{r, eval = F}
student_data <- data.frame(
  name = c("Alice", "Bob", "Charlie", "David", "Eve", "Frank"),
  score = c(85, 92, 78, NA, 95, 88)
)
student_data %>%
  na.omit() %>%
  transform(z_score = (score - mean(score)) / sd(score)) %>%
  subset(z_score > 0) %>%
  .[order(-.$z_score), ] # R is going to replace the two dots with the previously generated dataset
```

---

## `tidyr`

### **Key Functions in `tidyr`**

The `tidyr` package provides functions for reshaping and tidying data efficiently. Below is a summary of some of the most commonly used functions:

```{r echo=FALSE, results="asis"}
# Create the table as a data frame
tidyr_functions <- data.frame(
  Function = c("`pivot_longer()`", "`pivot_wider()`", "`separate()`", "`unite()`", "`fill()`", "`drop_na()`"),
  Purpose = c(
    "Convert wide data to long format",
    "Convert long data to wide format",
    "Split a column into multiple columns",
    "Merge multiple columns into one",
    "Fill in missing values",
    "Remove missing values"
  )
)

# Print the table using knitr::kable() for nice formatting
knitr::kable(tidyr_functions, caption = "Summary of Key `tidyr` Functions")
```

### Reshaping data

One of the most common tasks in `tidyr` is reshaping wide data into long format (and vice versa)
  - Long format: 
  each row corresponds to a unique combination of original a row id and a column name

```{r, eval = F}
wide_data <- data.frame(
  name = c("Alice", "Bob", "Charlie"), 
  math = c(90, 85, 78), 
  science = c(88, 92, 80)
)
# Wide -> long
long_data <- wide_data %>%
  pivot_longer(cols = c(math,science), names_to = "subject", values_to = "score")
long_data

# Long -> wide
wide_again <- long_data %>%
  pivot_wider(names_from = subject, values_from = score)
wide_again
```

### Splitting columns

```{r, eval = F}
data <- data.frame(full_name = c("Alice Johnson", "Bob Smith"))
# Split full_name into first_name and last_name
separated_data <- data %>%
  separate(full_name, into = c("first_name", "last_name"), sep = " ")
separated_data
```

### Merging columns

```{r, eval = F}
data <- data.frame(first_name = c("Alice", "Bob"), last_name = c("Johnson", "Smith"))
# Merge first_name and last_name into full_name
united_data <- data %>%
  unite(full_name, first_name, last_name, sep = "_")
united_data
```

### Removing rows with missing values

```{r, eval = F}
data <- data.frame(name = c("Alice", "Bob", "Charlie"),
                   score = c(90, NA, 85))
clean_data <- data %>% drop_na()
clean_data
```

### Filling missing values with neigbours

```{r, eval = F}
pets <- tibble::tribble(
  ~rank, ~pet_type, ~breed,
  1L,        NA,    "Boston Terrier",
  2L,        NA,    "Retrievers (Labrador)",
  3L,        NA,    "Retrievers (Golden)",
  4L,        NA,    NA,
  5L,        NA,    "Bulldogs",
  6L,     "Dog",    "Beagles",
  1L,        NA,    "Persian",
  2L,        NA,    "Maine Coon",
  3L,        NA,    "Ragdoll",
  4L,     "Cat",    "Exotic",
  5L,        NA,    "Siamese",
  6L,        NA,    "American Short"
)
pets_filled1 <- pets %>% fill(breed, .direction = "down")
pets_filled2 <- pets %>% fill(pet_type, .direction = "down")
```

### Filling missing values with customized values

```{r, eval = F}
pets$pet_type %>% replace_na('UnknownType')
pets_filled3 = pets %>% replace_na(list(pet_type = 'UnknownType', breed = 'UnknownBreed'))
```

---

## `dplyr`

### Why use `dplyr`

- Simplifies data manipulation (compared to base `R`)
- Readable and intuitive syntax (using verb-based functions)
- Faster execution (optimized with `C++`)
- Works seamlessly with `tibbles` and other `tidyverse` members
- Uses the pipe operator for chaining multiple operations

### Key functions in `dplyr`

| Function           | Purpose                                              |
|--------------------|------------------------------------------------------|
| `filter()`         | Select rows based on conditions                      |
| `slice()`          | Select specific rows by position                     |
| `select()`         | Choose specific columns                              |
| `mutate()`         | Create or modify columns                             |
| `summarize()`      | Compute summary statistics                           |
| `group_by()`       | Group data for aggregation                           |
| `arrange()`        | Sort rows by column values                           |
| `distinct()`       | Remove duplicate rows                                |
| `join()` functions | Merge datasets (`left_join()`, `inner_join()`, etc.) |

### Removing duplicate rows

```{r, eval = F}
students <- tibble(
  name = c("Alice", "Bob", "Charlie", "Alice", "Eve", "Charlie", "Eve"),
  sex = c("F", "M", "M", "F", "F", "M", "F"),
  subject = c("Math", "Math", "Science", "Math", "Science", "Science", "Science"),
  score = c(88, 92, 85, 88, 78, 85, 78)
)
students

# Removing exact duplicate rows
students %>% distinct()

# Removing duplicates based on one column
students %>% distinct(name, .keep_all = TRUE)

# Removing duplicates based on multiple columns
students %>% distinct(name, subject, .keep_all = TRUE)
```

### Selecting rows based on conditions

```{r, eval = F}
mtcars %>% filter(cyl == 6)  
# base R alternative: mtcars[mtcars$cyl == 6, ]
```

### Selecting rows by position

```{r, eval = F}
starwars %>% slice(2,4)  # Select the 2nd and 4th rows
# base R alternative: starwars[c(2,4), ]

starwars %>% slice_head(n = 3)  # Select the first 3 rows
# base R alternative: head(starwars, 3)

starwars %>% slice_tail(n = 2)  # Select the last 2 rows
# base R alternative: tail(starwars, 2)

set.seed(718)  # Ensure reproducibility
starwars %>% slice_sample(n = 3)  # Randomly select 3 rows
# base R alternative: starwars[sample(nrow(starwars), 3),]

starwars %>% slice_max(order_by = height, n = 2)  # Select top 2 highest characters
# base R alternative: starwars[order(-starwars$height), ][1:2, ]

starwars %>% slice_min(order_by = height, n = 2)  # Select top 2 lowest characters
# base R alternative: starwars[order(starwars$height), ][1:2, ]
```

### Selecting columns

```{r, eval = F}
mtcars %>% select(mpg, hp, wt)  
# base R alternative: subset(mtcars, select = c(mpg, hp, wt))

mtcars %>% select(-c(mpg, hp, wt))
mtcars %>% select(!c(mpg, hp, wt))
# base R alternative: subset(mtcars, select = -c(mpg, hp, wt))
```

### Creating new columns that are functions of existing variables

```{r, eval = F}
mtcars %>% mutate(weight_kg = wt * 453.5)  
# base R alternative: mtcars$weight_kg <- mtcars$wt*453.5
```

### Summarizing data

```{r, eval = F}
mtcars %>% summarize(avg_mpg = mean(mpg), max_hp = max(hp)) 
# base R alternative: data.frame(avg_mpg = mean(mtcars$mpg), max_hp = max(mtcars$hp))
```

### Grouping data by one or more variables before further move

```{r, eval = F}
mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg), max_hp = max(hp))
# (imperfect) base R alternative: 
# aggregate(cbind(mpg, hp) ~ cyl, data = mtcars, FUN = function(x) c(mean = mean(x), max = max(x)))

mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg), max_hp = max(hp))
# (imperfect) base R alternative: 
# aggregate(cbind(mpg, hp) ~ cyl+gear, data = mtcars, FUN = function(x) c(mean = mean(x), max = max(x)))

mtcars %>%
  group_by(cyl, gear) %>%
  count()
# (imperfect) base R alternative: 
# aggregate(mpg ~ cyl+gear, data = mtcars, FUN = length)

mtcars %>%
  group_by(cyl, gear) %>%
  mutate(avg_mpg_cy_gear = mean(mpg))
# (imperfect) base R alternative: 
# group_means <- aggregate(mpg ~ cyl + gear, data = mtcars, FUN = mean)
# colnames(group_means)[3] <- "avg_mpg_cy_gear"
# merge(mtcars, group_means, by = c("cyl", "gear"))
```

### Arranging rows

```{r, eval = F}
mtcars %>% arrange(mpg) # Sort by mpg in ascending order
# base R alternative: mtcars[order(mtcars$mpg), ]

mtcars %>% arrange(desc(mpg)) # Sort by mpg in descending order
# base R alternative: mtcars[order(-mtcars$mpg), ]

mtcars %>% arrange(cyl, desc(mpg)) # Sort by cyl (ascending), then by mpg (descending)
# base R alternative: mtcars[order(mtcars$cyl, -mtcars$mpg), ]

mtcars %>% arrange(rownames(mtcars)) # Sort by cars name in alphabetical order
# base R alternative: mtcars[order(rownames(mtcars)), ]
```

### Concatenating multiple functions

```{r, eval = F}
mtcars %>%
  filter(cyl == 6) %>%
  select(mpg, hp, wt) %>%
  arrange(desc(mpg))
# base R alternative:
# mtcars_filtered <- mtcars[mtcars$cyl == 6, ]
# mtcars_selected <- mtcars_filtered[, c("mpg", "hp", "wt")]
# mtcars_sorted <- mtcars_selected[order(-mtcars_selected$mpg), ]
```

### Merging datasets

```{r, eval = F}
# two data frames on student names and scores, respectively
p_names <- data.frame(
  patient_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David")
)
results <- data.frame(
  patient_id = c(2, 3, 5),
  exam_res = c(85, 90, 78)
)
p_names
results

# Keeps all rows from p_names, adding available results
# Use left_join() when your primary dataset is on the left
left_join(p_names, results, by = "patient_id")

# Keeps all rows from results, adding available p_names
# Use right_join() when your primary dataset is on the right
right_join(p_names, results, by = "patient_id")

# Keep only matched rows
# Use inner_join() when you only need matched records
inner_join(p_names, results, by = "patient_id")

# Keep all rows from both tables
# Use full_join() when you want to retain everything
full_join(p_names, results, by = "patient_id")
```

---

## `ggplot2`

### Why use `ggplot2`

- One of the most powerful and flexible visualization tools in `R`
- Building plots layer by layer
- Providing highly customizable and publication-quality visualizations

### Basic syntax of `ggplot2`

`ggplot2` follows a layered grammar of graphics, 
where each component adds a layer to the plot.

```{r, eval = F}
ggplot(
  data = <DATA>, # Initializes the plot with a dataset
  aes(
    x = <X-VAR>, y = <Y-VAR>, color = <COLOR-VAR>, # maps data variables to visual properties
    ...
  ) 
) +
  <GEOM_FUNCTION>() +  # geometric shapes (i.e., type of plot; e.g., points, bars and lines)
  <SCALE_FUNCTION>() + #  sizes and scales
  <FACET_FUNCTION>() + # creates small multiples
  <THEME_FUNCTION>() + # customizes the plot appearance
  <LABEL_FUNCTION>()   # labels and titles
```

### Scatter plot

A scatter plot visualizes the relationship between two numerical variables
(e.g., `mass` vs `height`).

```{r, eval = F}
ggplot(
  starwars, 
  aes(x = mass, y = height)
) +
  geom_point() + # scatter plot
  labs( # titles and axis labels
    title = "Height vs. Mass of Star Wars Characters",
    x = "Mass (kg)", y = "Height (cm)"
  ) +
  theme_minimal() # background theme
```

Filter out NAs and extreme values before plotting.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height) & !is.na(mass) & mass < 1000) %>%
  ggplot(
    aes(x = mass, y = height)
  ) +
  geom_point() +
  labs(
    title = "Height vs. Mass",
    x = "Mass (kg)", y = "Height (cm)"
  ) +
  theme_bw()
```

Fit a trend line with prediction intervals using lm, loess, etc

```{r, eval = F}
starwars %>% 
  filter(!is.na(height) & !is.na(mass) & mass < 1000) %>%
  ggplot(
    aes(x = mass, y = height)
  ) +
  geom_point() +
  labs(
    title = "Height vs. Mass",
    x = "Mass (kg)", y = "Height (cm)"
  ) +
  theme_bw()+
  geom_smooth(method = lm, se = T)
```

Group points and color them by a categorical variable (e.g., `sex`).

```{r, eval = F}
starwars %>% 
  filter(!is.na(height) & !is.na(mass) & mass < 1000) %>%
  replace_na(list(sex = 'Unknown')) %>%
  ggplot(
    aes(x = mass, y = height, color = sex)
  ) +
  geom_point() +
  # scale_color_manual( # Customize colors
  #   values = c("female" = "blue", "male" = "red", 'Unknown'='yellow')
  # ) + 
  labs(
    title = "Height vs. Mass",
    x = "Mass (kg)", y = "Height (cm)", color = "sex"
  ) +
  theme_light()
```

Differentiate points using `shape` or `size`.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height) & !is.na(mass) & mass < 1000) %>%
  replace_na(list(sex = 'Unknown')) %>%
  ggplot(
    aes(
      x = mass, y = height, 
      shape = sex,  # different symbols for different sexs
      size = mass # larger mass values create bigger points (called bubble plot sometimes)
    ) 
  ) +
  geom_point() +
  labs(
    title = "Height vs. Mass",
    x = "Mass (kg)", y = "Height (cm)", shape = "sex", size = 'Mass'
  ) +
  theme_grey()
```

Create subplots by a categorical variable.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height) & !is.na(mass) & mass < 1000) %>%
  replace_na(list(sex = 'Unknown')) %>%
  ggplot(aes(x = mass, y = height,color = sex)) +
  geom_point() +
  labs(
    title = "Height vs. Mass by sex",
    x = "Mass (kg)", y = "Height (cm)", size = 'Mass'
  ) +
  theme_minimal() +
  facet_wrap(~sex) +
  geom_smooth(method = lm, se= T)
```

### Jitter plot

A jitter plot reduces overplotting in scatter plots when points overlap.
  - Overplotting occurs when too many data points overlap, 
  making it difficult to distinguish individual values in a plot. 
  This happens especially in scatter plots 
  when multiple points share the same or very similar coordinates.
  For example, age is measured in years,
  and height is given in centimeters.
  If you construct a scatter plot of weight versus age for a sufficiently 
# large sample of people, there might be many people recorded 
# as, say, 29 years and 70 kg, and therefore many markers 
# plotted at the point (29, 70).

```{r, eval = F}
starwars %>% 
  filter(!is.na(height) & !is.na(mass) & mass < 1000) %>%
  ggplot(
    aes(x = mass, y = height, color = sex)
  ) +
  geom_point() +
  geom_jitter(
    width= .2, # each point is randomly moved up to 0.2 units left or right.
    height= .2, # each point is randomly moved up to 0.2 units up or down.
    alpha = 0.5, color = "purple" # the transparency and color for jittered points
  ) + 
  labs(
    title = "Height vs. Mass",
    x = "Mass (kg)", y = "Height (cm)"
  ) +
  theme_bw()
```

### Histogram

Count (number of observations per bin) as the y-axis.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height)) %>%
  ggplot(aes(x = height)) +
    geom_histogram(bins = 10, fill = "steelblue", color = "black") +
    labs(title = "Distribution of Character Heights",
         x = "Height (cm)", y = "Count") +
    theme_minimal()
```

Relative frequency (proportion of total observations per bin) as the y-axis.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height)) %>%
  ggplot(aes(x = height)) +
    geom_histogram(
      aes(y = after_stat(count / sum(count))),
      bins = 10, fill = "steelblue", color = "black"
    ) +
    labs(title = "Distribution of Character Heights",
         x = "Height (cm)", y = "Count") +
    theme_minimal()
```

Add a scaled fitted curve.

```{r, eval = F}
binwidth_value = 10
starwars %>% 
  filter(!is.na(height)) %>%
  ggplot(aes(x = height)) +
    geom_histogram(
      aes(y = after_stat(count / sum(count))),
      binwidth = binwidth_value, fill = "steelblue", color = "black"
    ) +
    labs(title = "Distribution of Character Heights",
         x = "Height (cm)", y = "Count") +
    theme_minimal()+
    geom_density(aes(y = after_stat(density)*binwidth_value), color = "red", linewidth = 1.2)
```

Density (=relative frequency divided by binwidth) as the y-axis with the fitted density curve.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height)) %>%
  ggplot(aes(x = height)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 20, fill = "steelblue", color = "black"
    ) +
    labs(title = "Distribution of Character Heights",
         x = "Height (cm)", y = "Count") +
    theme_minimal()+
    geom_density(aes(y = after_stat(density)), color = "red", linewidth = 1.2)
```

### Bar plot

Visualize categorical data by showing counts (by default)

```{r, eval = F}
starwars %>% 
  ggplot(aes(x = species)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Character Count by Species",
       x = "Species", y = "Count") +
  theme_minimal() +
  coord_flip()  # Flip for better readability
```

Visualize categorical data by showing proportions
  - Include `aes(y = after_stat(count / sum(count)))` in `geom_bar()`

```{r, eval = F}
starwars %>% 
  ggplot(aes(x = species)) +
  geom_bar(
    aes(y = after_stat(count / sum(count))),
    fill = "steelblue", color = "black") +
  labs(title = "Character Count by Species",
       x = "Species", y = "Proportion") +
  theme_minimal() +
  coord_flip()
```

Stacked bar chart

```{r, eval = F}
starwars %>% 
  ggplot(aes(x = fct_rev(species), fill = sex)) + # Reverses the order of species to correct flipping
  geom_bar(position = "stack") +  # position = "fill" or "stack"
  labs(title = "Stacked Bar Chart",
       x = "sex", y = "Count") +
  theme_minimal() +
  coord_flip()
```

### Polar chart

A polar chart is simply a bar chart 
with a single category that is transformed using `coord_polar(theta = "y")`.

```{r, eval = F}
ggplot(starwars, aes(x = '', fill = sex)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Numbers of Characters by sex", x='', y='') +
  theme_minimal()
```

Change the labels to proportions instead of counts.

```{r, eval = F}
starwars %>%
  count(sex) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = "", y = prop, fill = sex)) +
  geom_bar(stat = "identity", width = 1) + # stat = "count" (by default), "identity" (precomputed values)
  coord_polar(theta = "y") + 
  labs(title = "Proportion of Characters by sex", x='', y='') +
  theme_minimal()
```

### Box plot

Compare heights by sex.

```{r, eval = F}
starwars %>% 
  filter(!is.na(height)) %>%
  replace_na(list(sex = 'Unknown')) %>%
  ggplot(aes(x = sex, y = height, fill = sex)) + # "fill=sex": paint boxes by sex
  geom_boxplot() +
  labs(title = "Height Distribution by sex",
       x = "sex", y = "Height (cm)") +
  theme_minimal()
```

### Density plot

Shows distributions of continuous variables

```{r, eval = F}
starwars %>%
  filter(!is.na(mass) & !is.na(sex) & mass<1000) %>%
  ggplot(aes(x = mass, fill = sex)) +
  geom_density(alpha = 0.6) + # alpha controls the transparency
  labs(title = "Density Plot of Mass by sex",
       x = "Mass (kg)",
       y = "Density") +
  theme_minimal()
```

### Violin plot with a box plot inside

A combination of a box plot and a density plot.

| Component | Description |
|---|---|
| Violin Shape | Density curve. |
| Thick Line | Sample median the dataset. |
| Quartiles (Box Edges)| 25% quantile (Q1) and 75% quantile (Q3). |
| Whiskers (Thin Lines) | Extend up to 1.5x(Q3-Q1). |
| Outliers | Points outside the whiskers. |

```{r, eval = F}
starwars %>%
  ggplot(aes(x = sex, y = height, fill = sex)) +
  geom_violin(alpha = 0.5) + 
  geom_boxplot(width = 0.1) +  
  labs(title = "Violin Plot with Box Plot",
       x = "sex", y = "Height (cm)") +
  theme_minimal()
```

### Contour plot 

Visualize a 3D surface in 2D by drawing lines of equal values.

```{r, eval = F}
x <- seq(-pi, pi, length = 50)
y <- x
z_matrix <- outer(x, y, function(x, y) x^2 + y^2)
df <- expand.grid(x = x, y = y) %>% mutate(z = as.vector(z_matrix))
ggplot(df, aes(x = x, y = y, z = z)) +
  geom_contour(
    bins = 10, # number of contour levels
    color = "blue"
  ) +
  metR::geom_text_contour(skip = 0, stroke = 0.2) + # Add contour labels
  labs(title = "Function Visualization: f(x, y) = x^2 + y^2",
       x = "X-axis", y = "Y-axis") +
  theme_minimal()
```

Could be more illustrative by filling the regions between adjacent lines.

```{r, eval = F}
ggplot(df, aes(x = x, y = y, z = z)) +
  geom_contour_filled(bins = 10) +
  scale_fill_viridis_d(direction = -1, option = "H") +
  labs(title = "Function Visualization: f(x, y) = x^2 + y^2",
       x = "X-axis", y = "Y-axis", fill = "Z value") +
  theme_minimal()
```

Interactive 2D contour plot with `plotly`

```{r, eval = F}
p_contour = ggplot(df, aes(x = x, y = y, z = z)) +
  geom_contour(
    bins = 10, # number of contour levels
    color = "blue"
  ) +
  labs(title = "Function Visualization: f(x, y) = x^2 + y^2",
       x = "X-axis", y = "Y-axis") +
  theme_minimal()
plotly::ggplotly(p_contour)
```

Interactive 3D plot with `plotly`

```{r, eval = F}
library(plotly)
plot_ly(x = ~x, y = ~y, z = ~z_matrix , type = "surface", colorscale = "Viridis") %>%
  layout(
    title = "3D Surface Plot: f(x, y) = x^2 + y^2",
    scene = list(
      xaxis = list(title = "X-axis"),
      yaxis = list(title = "Y-axis"),
      zaxis = list(title = "Z-axis")
    )
  )
```

### Heatmap

Representing the magnitude of values as colors.

```{r, eval = F}
df %>%
  ggplot(aes(x = x, y = y, fill = z)) +
  geom_tile() +
  scale_fill_gradient(low="yellow", high="red") +
  labs(title = "Function Visualization: f(x, y) = x^2 + y^2",
       x = "X-axis", y = "Y-axis", fill = "f(x,y)") +
  theme_minimal()
```

### Line plot

Showing relationships between two variables.

```{r, eval = F}
economics %>%
  ggplot(aes(x = date, y = unemploy)) +
  geom_line(color = "blue", linewidth = 1) +
  labs(title = "U.S. Unemployment Over Time",
       x = "Time", y = "Number of unemployed in thousands") +
  theme_minimal()
```

Two-line plot.

```{r, eval = F}
economics %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = unemploy, color = "Unemployment"), linewidth = 1) +
  geom_line(aes(y = pop, color = "total population"), linewidth = 1) +
  scale_color_manual(values = c("Unemployment" = "blue", "total population" = "red")) +
  labs(title = "Total Population and Unemployment Over Time",
       x = "Time", y = "Number in thousands", color = "Variable") +
  theme_minimal()
```

Two trajectories on different scales with a second y-axis added.

```{r, eval = F}
economics %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = unemploy, color = "Unemployment"), linewidth = 1) +  
  geom_line(aes(y = psavert*1000, color = "Personal Savings Rate"), linewidth = 1) +  
  scale_color_manual(values = c("Unemployment" = "blue", "Personal Savings Rate" = "red")) +
  scale_y_continuous(sec.axis = sec_axis(~./1000, name = "Savings Rate (%)")) +
  labs(title = "U.S. Unemployment and Savings Rate Over Time",
       x = "Year", y = "Unemployed People (in Thousands)", color = "Variable") +
  theme_minimal()
```

### Saving figures

```{r, eval = F}
ggsave(filename="first_ggplot.png")  ## Save the last plot
g = starwars %>% 
  filter(!is.na(height)) %>%
  ggplot(aes(x = height)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 20, fill = "steelblue", color = "black"
    ) +
    labs(title = "Distribution of Character Heights",
         x = "Height (cm)", y = "Count") +
    theme_minimal()+
    geom_density(aes(y = after_stat(density)), color = "red", linewidth = 1.2)
g
ggsave(g, filename="second_ggplot.pdf") ## Save a specific plot
```


